Rodent trapping studies can improve our understanding of the hazards of zoonotic pathogen spillover and novel zoonotic pathogen emergence.

### 2022-05-11 reports/manuscript_v4.Rmd
updating scripts

---
title: "Synthesising rodent trapping studies in West Africa to understand the hazard of zoonotic infectious disease outbreaks."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  word_document
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: elsevier_vancouver.csl
---

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
source(here::here("scripts", "manuscript_library.R"))
```

@@

**To complete after written**

@@

Zoonotic infectious disease spillover events into human populations  (animal-to-human transmission) can have important health and economic consequences at both the global and local level (e.g. SARS-CoV-2 and Ebola virus disease) (reference). The number of zoonotic disease spillover events and the emergence of novel zoonotic pathogens are projected to increase under intensifying anthropenic pressure driven by increased human populations, urbanisation, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. West Africa has been identified as a nexus for these processes and a hotspot for both the risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence and [@grace_mapping_2012; @han_global_2016; @allen_global_2017]. Further, two taxa - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_global_2016]. Specifically, West Africa is identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. In West Africa rodents are involved in the transmission of multiple zoonotic diseases, including, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021]. Traits associated with "fast" life histories are prevalent in species that thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].

Despite the importance of understanding these complex systems in the hazard of zoonotic spillover, current evidence on host-pathogen associations is affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are based on global datasets of host species distributions and host-pathogen associations (e.g., IUCN, GBIF, GIDEON) that suffer from several biases [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations), resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching global datasets and counteracting inherent biases [@bovendorp_optimising_2017]. Rodent trapping provides a useful method to describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in global datasets. Studies have been conducted in West Africa to investigate the distribution of rodent speciesm, their species assemblages, the prevalence of known zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021]. However, data from these studies have not been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.

Here, we synthesise the rodent trapping studies conducted across West African from a search of literature between 1964-2019. First, we investigate the geographic biases in the rodent trapping dataset in relation to human population density and land use classification. Second, we compare this to global host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions. Third, we compare identified host-pathogen associations with global datasets (CLOVER) to understand discrepencies in rodent host-pathogen associations and report the proportion positive for each pathogen. Finally, within our dataset we investigate host-pathogen spatial extent to identify areas of undersampling of pathogens within host ranges. We expect that rodent trapping studies will provide an important additional source of high resolution data that can be used to enrich global datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa. We find that rodent trapping studies identify greater geographic ranges of important rodent hosts of zoonotic diseases than are reported from global datasets while also identifying locations and habitats in which further sampling is required. We also identify host-pathogen associations that are not reported in global host-pathogen association data, however, there remains systematic undersampling for these pathogens across the host ranges.

@@

We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary 1.). Data was abstracted into a Google Sheets document, which was archived on completion of data extraction (Supplementary 2.). Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest. We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.

@@

### Environmental covariates

### Rodent species ranges

### Rodent presence

@@

CLOVER a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw 2020) was released in 2021 and represents the most comprehensive global repository of host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi [clover_paper]. Data were obtained from an archived version of the dataset [clover_dataset].

@@

We identified 4,282 relevant citations, with 126 rodent trapping studies included. The earliest trapping studies were conducted in 1964, with increasing numbers of studies being performed annually. The median year of first trapping activity was 2005, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary figure 1.). Studies were conducted in 14 West African countries with no studies reported from Gambia or Togo, at 1,331 sites (Figure 1A.). A minority (31, 25%) of studies trapped at a single study site, with 46 (37%) trapping at between two and five sites, the remaining 49 studies (38%) trapped at between six and 93 study sites. There were 581,426 reported trap nights, with 384,983 trap nights imputed for studies with no recording of trapping effort (Figure 1B.).

```{r trap-locations-trapnights, out.width = "120%", fig.cap = "Figure 1. A) The location of trapping sites in West Africa. No sites were recorded from Togo or Gambia. There is important heterogeneity observed in the coverage of each country by trap night (colour) and location of sites. For example Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso. B) The distribution of trap nights performed at each study site on a log scale with a median of 244 trap nights (IQR 115-494) performed at each site"}

knitr::include_graphics(here("figures", "Figure_1_updated.png"))

Within the included studies trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared to other habitats (3%, IQR 1-9%). In total 73,164 small mammals were trapped with 62,574 (85%) identified to species level and 7,760 (11%) identified to genus. The majority of the 147 identified species were Rodentia (112, 76%) of which Muridae (82, 73%) were the most common family.

62 studies tested for 32 microorganisms, defined to species or genus leve that are known or potential pathogens. The majority of studies tested for a single microorganism (39, 62%). The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (27, 43%), *Borrelia sp.* (8, 13%), *Bartonella sp.* (4, 6%) and *Toxoplasma gondii* (4, 6%). The majority of studies used Polymerase Chain Reaction to detect microorganisms (32, 52%), with fewer studies using molecular based tests (11, 18%) or histological or direct visualisation assays (8, 13%).

From 32,014 individual rodent samples we produced 339 host-pathogen pairs. With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *A. niloticus* being assayed for at least 18 microorganisms.

@@

We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative regions in West Africa. The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments due to important differences in trap success between these different environments. The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites where there was no record of the number of trap nights. We summarised the habitat types of trap sites based on information reported in the study; the dictionary used to match habitat types is available in Supplementary 3.

To better understand the potential bias associated with studies for zoonotic disease spillover or novel zoonotic pathogen emergence we limited further analyses to zoonotic disease studies only. We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association between number of trap nights, proportion of urban, cropland and forest land classification and human population density [@pedersen_hierarchical_2019]. Covariates were obtained for 2005, the median year studies were conducted, proportions of habitat types were obtained from Copernicus 2005 land cover classification, global population density estimates were obtained from SEDAC [@socioeconomic_data_and_applications_center_gridded_2021]. We performed sensitivity analysis by removing sites with imputed trapping effort. The model structure was specified as:

$$
\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie(log(Population density) + (Urban) + (Cropland) + (Shrubland) + (Forest) + (Longitude * Latitude))}
\end{equation}
$$

The model was used to identify locations of high and low trapping effort relative to population density and land use classification. Predictions were limited to regions around observed trap sites.

@@

We assessed the taxonomic coverage of the trapped taxa and for the seven species (*M. natalensis*, *R. rattus*, *M. erythroleucus*, *M. musculus*, *Arvicanthis niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) with the most trap locations we mapped the detection and non-detection of rodent species and compared this to detection data obtained from GBIF and species ranges from IUCN [@gbif_the_global_biodiversity_information_facility_occurrence_2021; @iucn_iucn_2021]. We identified rodent locations from GBIF and the rodent trapping studies within and outside the IUCN expected range. We located these to level 2 administrative regions to calculate the discrepencies between observed detections and non-detections with IUCN distributions.

@@

For studies investigating rodent zoonoses we produced host-pathogen pairs. For each host-pathogen pair we compared our records with those available in CLOVER reporting concordance and discordance. We identify host-pathogen pairs that are not present in CLOVER and contain assay results consistent with acute or prior infection.

@@

For the four pathogens sampled at the most locations (Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) we identify their host-pathogen associations. We identified the five species with most spatially complete sampling and associated these sampling locations with level 2 administrative regions. We compared the area of these regions to their IUCN range and detection range to calculate sampling coverage.

@@

Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in 14 West African nations. The areas with highest trap night density included the capital and large cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), Ghana (Accra),  Côte d'Ivoire (Abidjan), and Benin (Cotonou). Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 1A.).

Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats. However, urban habitats were undersampled.

Trapping effort, measured through trap nights, was biased towards areas with higher population density and predominantly urban and cropland land classifications (Figure 2.). In particular Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria had relatively high trapping efforts while South East Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger, and Burkina Faso were trapped at lower trapping effort than would be expected. Excluding sites with imputed trap nights reduced the number of trapping sites contributing data. Despite this limitation, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher than expected rates with Nigeria as being trapped at lower than expected rates (Supplementary Figure 2.).

```{r gam-trapping-effort, out.width = "120%", fig.cap = "Figure 2. Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion cropland cover, proportion forest cover and proportion urban habitat. Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor. Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort"}

knitr::include_graphics(here("figures", "Figure_2_updated.png"))

@@

For our species of interest, except *C. gambianus*, trapping studies provided more locations of detection and non-detection than were available from GBIF. For the West African endemic species IUCN ranges showed generally good concordance to both trapping studies and GBIF, although the range of *A. niloticus* and *P. daltoni* are greater than expected. The invasive species *R. rattus* and *M. musculus* are present in a much larger range than is expected (Figure 3A. and 3B.).

```{r comparing-trap-locations-A, out.width = "120%", fig.cap = "Figure 3A. Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range. R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges. No IUCN range for *M. musculus* is available for the study region."}

knitr::include_graphics(here("figures", "Figure_3_test_a.png"))

```{r comparing-trap-locations-B, out.width = "120%", fig.cap = "Figure 3B. Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range. R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges."}

knitr::include_graphics(here("figures", "Figure_3_test_b.png"))

We identified the proportion of a rodents expected range in which detections and non-detections occurred (Table 1.). GBIF sampling occurred over > 15% of a rodents expected range with detections occuring outside of IUCN ranges for all species, most noticeably for *A. niloticus* and *P. daltoni*. Trapping studies occurred over > 13% for all species except *C. gambianus* with detections occurring outside of IUCN ranges most noticeably for *A. niloticus* and *R. rattus*. Combining GBIF and trapping data increased coverage by a mean of 9% of the species range suggesting limited overlap between the regions studied. Non-detections of rodents occurred across large areas of their ranges with *A. niloticus* and *M. erythroleucus* not being detected in >15% of their expected range.

@@

table_1 <- read_rds(here("tables", "Table_1_updated.rds"))

@@

When limited to confirmed pathogens and classification of both pathogen and host to species level we identified 21 host-pathogen pairs among 11 rodent and 6 pathogen species (Figure 4.). We identified negative associations for 41 host-pathogen pairs among 31 rodent and 7 pathogen species. CLOVER contained 8 (38%) of our identified host-pathogen associations and positive associations for 6 (15%) of the negative associations.

```{r comparing-host-pathogen-interactions, out.width = "120%", fig.cap = "Figure 4. L) Identified species level host-pathogen associations through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. R) Identified species level host-pathogen associations through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset."}

knitr::include_graphics(here("figures", "Figure_4_updated.png"))

>>>>

---
title: "Rodent trapping studies can improve our understanding of the hazards of zoonotic pathogen spillover and novel zoonotic pathogen emergence."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  officedown::rdocx_document:
    tables:
      caption:
       style: false
    plots:
      caption:
        style: false
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: elsevier_vancouver.csl
---

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      crop = knitr::hook_pdfcrop)
source(here::here("scripts", "manuscript_library.R"))
```

@@

Rodents are important reservoirs of known zoonotic infectious diseases and novel zoonotic pathogens. Rodent reservoirs are globally distributed with West Africa containing multiple host species of zoonotic infectious diseases. Ongoing anthropogenic land use change alters the composition of rodent species assemblages which can alter the hazard of zoonotic disease spillover and emergence. Current global datasets to understand these hazards suffer from biases. Here, we synthesise data from 127 rodent trapping studies including over 72,000 individual rodent detections to understand the range and presence of important rodent reservoir species to quantify the bias in global datasets. We identify the rodent trapping studies are biased towards human dominated landscapes with important sampling biases across West Africa. We find that these rodent trapping studies can complement current global datasets on rodent population distributions which can minimise some of these observed biases. For four regionally important zoonotic infectious diseases we identify host-pathogen associations not included in a comprehensive global dataset. Rodent trapping studies can further enrich these resources through geolocating these associations and producing measures of pathogen prevalence within hosts. Finally, we find that for most of these diseases the proportion of a rodent hosts range that has been sampled remains small. Further rodent trapping studies are required to understand the true hazard of zoonotic spillover of these diseases across West Africa. Our results highlight that understanding rodent sampling bias is important when assessing the risk for zoonotic spillover events from rodents. The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data. We anticipate that this dataset can support the production of more complete understanding of the hazard of zoonotic spillover events.

@@

Zoonotic infectious disease spillover events into human populations (animal-to-human transmission) can have important health and economic consequences at both the global and regional level (e.g. SARS-CoV-2 and Ebola virus disease) [@judson_zoonoses_2021]. The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropenic pressure driven by increased human populations, urbanisation, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. West Africa is a nexus for these processes and a hotspot for both the risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@grace_mapping_2012; @han_global_2016; @allen_global_2017]. Two taxa - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_global_2016]. Specifically, West Africa has been identified as region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. Currently within West Africa rodents are involved in the transmission of multiple zoonotic diseases with large burdens on human health, including, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021]. Rodent species with "fast" life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].

Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are based on global datasets of host species distributions and host-pathogen associations (e.g., IUCN, GBIF, GIDEON) which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations), resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching global datasets and mitigating against their inherent biases [@bovendorp_optimising_2017]. Rodent trapping provides a useful method to describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in global datasets. Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of known zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021]. However, data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.

Here, we synthesise rodent trapping studies conducted across West African from a search of literature between 1964-2019. First, we investigate the geographic biases in the rodent trapping dataset in relation to human population density and land use classification. Second, we compare this to global host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions. Third, we compare identified host-pathogen associations with global datasets (CLOVER) to understand discrepencies in rodent host-pathogen associations and report the proportion positive for pathogens of interest. Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of undersampling of pathogens within their host ranges. We expect that rodent trapping studies will provide an important additional source of high-resolution data that can be used to enrich global datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa. We find that rodent trapping studies identify greater geographic ranges of important rodent hosts of zoonotic diseases than are reported from global datasets while also identifying locations and habitats in which further sampling is required. We also identify host-pathogen associations that are not reported in global host-pathogen association data, while identifying systematic undersampling for these pathogens across their host ranges.

@@

We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.). Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo]. Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest. We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.

@@

### Spatial covariates

Boundaries of countries and level-2 administrative regions were obtained from GADM 4.0.4 for the study region [@database_of_global_administrative_areas_gadm_2022]. Land cover classification in 2005, the median year in which trapping activity occurred, was obtained from Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022]. Proportion of land cover type within a level-2 administrative region was calculated. Human population density for 2005 was obtained from SEDAC at ~ 1km resolution, with median population density within each region extracted [@socioeconomic_data_and_applications_center_gridded_2021].

### Rodent species ranges and presence

Rodent species distributions were obtained as shapefiles from the IUCN redlist, shapefiles were cropped to the study region [@iucn_iucn_2021]. Rodent presence locations were obtained from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021].

@@

CLOVER a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw 2020) was released in 2021 and represents a comprehensive global repository of host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi [@gibb_data_2021]. Data were obtained from an archived version of the dataset [@gibb_viralemergenceclover_2021]. The dataset was limited to species trapped within our included studies, no spatial subsetting was possible.

@@

We identified 4,282 relevant citations, with 126 rodent trapping studies included (Supplementary Table 2.). The earliest trapping studies were conducted in 1964, with increasing numbers of studies being performed annually. The median year of first trapping activity was 2005, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.). Studies were conducted in 14 West African countries with no studies reported from Gambia or Togo, at 1,331 sites (Figure 1A.). A minority (31, 25%) of studies trapped at a single study site, with 46 (37%) trapping at between two and five sites, the remaining 49 studies (38%) trapped at between six and 93 study sites. There were 581,426 reported trap nights, with 384,983 trap nights imputed for studies with no recording of trapping effort (Figure 1B.).

![Figure 1: A) The location of trapping sites in West Africa. No sites were recorded from Togo or Gambia. There is important heterogeneity observed in the coverage of each country by trap night (colour) and location of sites. For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso. B) The distribution of trap nights performed at each study site on a log scale with a median of 244 trap nights (IQR 115-494) performed at each site](`r here("figures", "Figure_1_updated.png")`)

Within the included studies trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared to other habitats (3%, IQR 1-9%). In total 73,164 small mammals were trapped with 62,574 (85%) identified to species level and 7,760 (11%) identified to genus. The majority of the 140 identified species were Rodentia (110, 76%) and Soricomorpha (30, 21%), of which Muridae (82, 73%) were the most common family.

62 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens. Most studies tested for a single microorganism (39, 62%). The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (27, 43%), *Borrelia sp.* (8, 13%), *Bartonella sp.* (4, 6%) and *Toxoplasma gondii* (4, 6%). Most studies used Polymerase Chain Reaction to detect microorganisms (32, 52%), with fewer studies using molecular based tests (11, 18%) or histological or direct visualisation assays (8, 13%). From 32,014 individual rodent samples we produced 339 host-pathogen pairs. With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.

@@

We used the GPS coordinates of trapping locations and the number of trap nights to calculate trapping effort (trap night density) within level-2 administrative regions in West Africa. We imputed trap nights for studies with missing records using the number of trapped rodents and median trap success rate from studies reporting trapping effort stratified by habitat type. We calculated the proportion of land cover classes within each of these level-2 administrative, comparing surveyed regions to non-surveyed regions.

To better understand the potential bias associated with studies for zoonotic disease spillover or novel zoonotic pathogen emergence we limited further analyses to zoonotic pathogen studies only. We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association between number of trap nights, proportion of urban, cropland and forest land classification and human population density [@pedersen_hierarchical_2019]. The model was constructed in *mgcv* using the R statistical programme [@wood_generalized_2017; @r_core_team_r_2020]. Selection of the most parsimonious model was based on the adjusted-R^2^ and Deviance explained for each model. We performed sensitivity analysis by removing sites with imputed trapping effort. The final selected model structure was specified as:

$$
\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie((Urban) + (Forest) + (Longitude * Latitude))}
\end{equation}
$$

The model identified locations of high and low trapping effort relative to land use classification, with proportion urban land very highly correlated with human population density (*r* = 0.835, 95% C.I. 0.82-0.85, *p* < 0.001). Model inference was limited to regions around observed trap sites.

@@

We assessed the taxonomic coverage of the trapped taxa and for the seven species with the most trap locations (*M. natalensis*, *R. rattus*, *M. erythroleucus*, *M. musculus*, *A. niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped the detection and non-detection of rodent species and compared this to detection data obtained from GBIF and species ranges from IUCN [@gbif_the_global_biodiversity_information_facility_occurrence_2021; @iucn_iucn_2021]. We identified rodent locations from GBIF and the rodent trapping studies within and outside the IUCN expected range. We located these to level-2 administrative regions to calculate the discrepencies between observed detections and non-detections with IUCN distributions.

@@

For studies investigating rodent zoonoses we produced host-pathogen pairs. For each host-pathogen pair we compared our records with those available in CLOVER, reporting concordance and discordance. For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER. We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.

@@

For the four pathogens sampled at the most locations (Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) we identify their potential host species. For the five species with most spatially complete sampling we associate their sampling locations with level-2 administrative regions and compare this to their IUCN range and rodent trapping detection range to calculate a measure sampling coverage.

@@

Trap sites were situated in 273 of 1,414 (19.3%) level-2 administrative regions in 14 West African nations. The areas with highest trap night density included the capital and large cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), Ghana (Accra),  Côte d'Ivoire (Abidjan), and Benin (Cotonou). Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 1A.).

Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats. However, urban habitats were comparitively undersampled (Supplementary Figure 2.).

Trapping effort, measured through trap nights, was biased with increased effort in areas with higher proportion of urban land use and no-significant association with proportion of forest cover. The spatial term shows trapping bias towards increased effort in Northern Senegal, Eastern Sierra Leone, Eastern Guinea and coastal regions of Nigeria, Benin and Ghana, in contrast South-East Nigeria, Northern Nigeria and Liberia had relatively lower trapping effort (Figure 2). The most parsimonious GAM model (adjusted R^2^ = 0.523, Deviance explained = 61%) included terms for coordinates, proportion urban coverage and proportion tree coverage, model selection is shown in Supplementary Tables 3.1-3.4. In sensitivity analysis excluding sites with imputed trap nights reduced contributing data, however, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates with Nigeria being trapped at lower than expected rates (Supplementary Figure 3.).

![Figure 2. Relative trapping effort bias across West Africa from included studies adjusted for proportion urban land cover and proportion tree cover. Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor. Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Figure_2_updated.png")`)

@@

For our species of interest, except *C. gambianus*, trapping studies provided more distinct locations of detection and non-detection than were available from GBIF. For the West African endemic species IUCN ranges showed generally good concordance to both trapping studies and GBIF, although the range of *A. niloticus* and *P. daltoni* are greater than expected. The invasive species *R. rattus* and *M. musculus* are present in a much larger range than is expected (Figure 3A. and 3B.).

![Figure 3A.  Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area). R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges. M. musculus has no range in West African in IUCN data.](`r here("figures", "Figure_3_test_a.png")`)

![Figure 3B. Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area). R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges.](`r here("figures", "Figure_3_test_b.png")`)

We identified the proportion of a rodent's expected range in which detections and non-detections occurred (Table 1.). GBIF sampling occurred over > 15% of a rodent's expected range with detections occurring outside of IUCN ranges for all species, most noticeably for *A. niloticus* and *P. daltoni*. Trapping studies occurred over > 13% for all species except *C. gambianus* with detections occurring outside of IUCN ranges most noticeably for *A. niloticus* and *R. rattus*. Combining GBIF and trapping data increased coverage by a mean of 9% of the species range suggesting limited overlap between the regions studied. Non-detections of rodents occurred across large areas of their ranges with *A. niloticus* and *M. erythroleucus* not being detected in >15% of their expected range, suggestive of substantial heterogeneity fo presence across their range.

@@

table_1 <- read_rds(here("tables", "Table_1_updated.rds")) %>%
  set_caption(caption = "Table 1: Comparison of IUCN, GBIF and rodent trapping ranges for the 7 species trapped at the most sites.") %>%
  autofit()

@@

When limited to confirmed pathogens and classification of both pathogen and host to species level we identified 21 host-pathogen pairs among 11 rodent and 6 pathogen species (Figure 4.). We identified negative associations, through entirely negative assays, for 41 host-pathogen pairs among 31 rodent and 7 pathogen species. CLOVER contained 8 (38%) of our identified host-pathogen associations and positive associations for 6 (15%) of the negative associations.

![Figure 4. L) Identified species level host-pathogen associations through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. R) Identified species level host-pathogen associations through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4_updated.png")`)

CLOVER included 399 host-pathogen associations we do not observe in rodent trapping studies among our included rodent species. The majority of these 329 (82%) pairs are from species with global distributions (*M. musculus*, *R. rattus* and *R. norvegicus*), or from those with wide ranging distributions in sub-Saharan Africa (31, 9%) (i.e., *A. niloticus*, *M. natalensis* and *Atelerix albiventris*).

For pathogens not identified to species level (i.e. family or higher taxa only) we identified 128 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary Figure 4.), with CLOVER containing 52 (41%) of these associations.
